---
title: "Probing the Depths of Probit Regression"
author: "Bertrand Wilden"
date: "2023-06-21"
categories: [Bayes, brms, Tutorial, R]
max-description-length: 20
---

```{css, echo=FALSE}
.title {
  color: #f5f5f5;
}
```

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Packages Loaded"
library(tidyverse)
library(brms)
library(tidybayes)
library(marginaleffects)
library(unvotes)
library(ggdist)

theme_set(theme_minimal())
```

When do you use a probit model in statistics? When you have some data and want to probe it for answers! 

It is also a method for modeling a data generating process which results in a binary 0/1 outcome. Maybe you are trying to explain whether someone votes in a particular election, whether a web user clicks on a link, or whether writing statistics blog posts helps its author get a job. Probit models are closely related to their more popular cousin: the logistic, or logit regression. So closely related, in fact, that I can't think of any decisive reason why someone would choose one over the other. I like probit regression because I always aspire to maintain a high level of statistical probity in my work. In this post I am going to explain what a probit model is, how to fit one in R, and introduce a brand new nested hurdle probit model.

## What is a Probit Model?

As I mentioned before, we can use a probit model when the outcome of interest is some binary variable. Binary outcomes arise out of what's known as a *Bernoulli* distribution, which we write as:

$$
y_i \sim \text{Bernoulli}(p_i)
$$

The $y_i$ above stands for the observed 1's and 0's in our data, and the $p_i$ is the probability of a particular $y_i$ equaling 1. For example, the heads (1) and tails (0) we observe from flipping a coin repeatedly would be generated from a Bernoulli distribution with $p_i = 0.5$. We know that $p_i = 0.5$ in the coin flipping example because we have no information that could lead us to expect one outcome over another---hence a 50/50 probability of getting heads. For more complicated data generating processes, however, we don't know $p_i$ in advance. Instead, we use statistical models to rank the relative plausibility of every possible value of $p_i$ based on the data we have collected. The general term for this sort of procedure is called Bayesian updating.

How do we go about constructing plausibility rankings for $p_i$? This is where our probit (*prob*ability un*it*) model comes in. Say we have a number of observed variables, $X_{1i}, X_{2i}, X_{3i}$ whose linear combination we think affects whether $y_i = 1$ or $y_i = 0$ (in other words, $p_i$). In this case we would like to use an equation that looks like:

$$
p_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \beta_3X_{3i}
$$

Here the $\beta$ terms represent the marginal effect each of the $X$ variables has on $p_i$. But wait! We defined $p_i$ earlier as a probability value---a real number between 0 and 1. The Bernoulli distribution can't give us values for $y_i$ if $p_i$ is not a valid probability. There is nothing in the equation above that enforces the value of the right hand side, $\beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \beta_3X_{3i}$ to be between 0 and 1 🙁. Luckily there is a little guy known as $\Phi$ who is ready to come to our rescue. We simply wrap our right-hand expression in $\Phi$'s loving embrace and it takes care of transforming the value of these linear predictors to a value on the probability scale of $p_i$:

$$
p_i = \Phi(\beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \beta_3X_{3i})
$$

What is $\Phi$ and how does it work? It represents the cumulative distribution function (CDF) for a standard Normal distribution, which means that it takes as input any real number and gives you back the probability of getting that value or less out of a standard Normal distribution.[^1] The graph below shows what this function looks like.

[^1]: The standard Normal distribution is a Normal distribution with mean 0 and standard deviation 1.

```{r}
ggplot() +
  stat_function(fun = pnorm, # pnorm is the Normal CDF function in R
                xlim = c(-4, 4)) +
  geom_segment(aes(x = 1.5, xend = 1.5, y = 0, yend = pnorm(1.5)),
                   color = "red", linetype = "dashed") +
  geom_segment(aes(x = -4, xend = 1.5, y = pnorm(1.5), yend = pnorm(1.5)),
                   color = "red", linetype = "dashed") +
  scale_x_continuous(breaks = seq(-4, 4, by = .5)) +
  scale_y_continuous(breaks = seq(0, 1, by = .1)) +
  labs(title = "Standard Normal CDF",
       x = "X",
       y = "Cumulative Probability")
```

If we supply $\Phi$ with an input of 1.5, it gives us an output of `r round(pnorm(1.5), 2)`.

The standard Normal CDF's ability to turn any real number into a probability between 0 and 1 is what allows us to build a model for $p_i$, which then tells us something about the Bernoulli distributions which produced our binary outcome variable $y_i$. As an aside, the difference between probit and logit models comes down to how they transform the linear predictor to the probability scale. Probit uses $\Phi$ and logit uses the log-odds transformation.[^2] This changes the scale of the $\beta$ parameters, and therefore how to interpret their values plus informing us about what prior values we should assign, but otherwise the two models are the same.

[^2]: $\text{ln}\left(\frac{x}{1-x}\right)$

## The Probit Model in Action

Let's jump into some real data to check out the probit model in action. We're going to use data from the [United Nations General Assembly Voting Data R package](https://github.com/dgrtwo/unvotes)[^3] which contains information about how each country voted on particular UN resolutions. Our goal is to build a model which predicts how the United States votes on each resolution.

[^3]: Erik Voeten "Data and Analyses of Voting in the UN General Assembly" Routledge Handbook of International Organization, edited by Bob Reinalda (published May 27, 2013)

```{r}
#| warning: false
un <- un_votes |> 
  left_join(un_roll_call_issues, by = "rcid") |> 
  left_join(un_roll_calls, by = "rcid") |> 
  mutate(vote = case_when(vote == "yes" ~ 1,
                          vote == "no" ~ 0,
                          vote == "abstain" ~ 2))
```

The first step will be to join the three data sets in the **unvotes** package together so that we have all the possible variables at our disposal. In order to make our lives easier during the modeling steps we need to recode the outcome variable, `vote`, such that it takes numeric values: "yes" = 1, "no" = 0, and "abstain" = 2. The basic probit model can only handle binary outcome variables, so we will ignore abstentions for the time being.

```{r}
#| warning: false
un <- un |> 
  group_by(rcid) |> 
  mutate(ussr_vote = ifelse(any(country_code == "RU" & vote == 1), 1, 0)) |> 
  ungroup() |> 
  mutate(issue = ifelse(is.na(issue), "Other", as.character(issue))) |> 
  filter(country_code == "US",
         date < as.Date("1991-12-26"))
```

Next, let's assemble our predictor variables. Now I'm no international relations expert, but I conjecture that one of the best ways to predict how the US will vote on a resolution is by looking at how its longtime foe the USSR voted. Naturally this limits our analysis to the time period before Christmas day 1991---which calls into question our use of the word "prediction". What does it mean to predict events which took place over 30 years ago? Let's overlook this detail for the sake of exposition.

In addition to a variable for how the USSR voted on resolutions, we will include a variable in the model for the issue, or topic, of each resolution. I created a new category called "Other" which contains resolutions with missing `NA` issues so that these observations are still included in the analysis.

```{r}
#| warning: false
un |> 
  janitor::tabyl(issue)
```

Now we can finally fit our model! This is a very simple probit regression so we can use the [brms R package](https://paul-buerkner.github.io/brms/). Because all our predictor variables are categorical, we'll use the `0 + ...` formula syntax for [index coding](https://bookdown.org/content/4857/the-many-variables-the-spurious-waffles.html#categorical-variables). The code `prior(normal(0, 2), class = b)` sets weakly informative priors for the coefficients on all these variables (notice how on the standard Normal CDF graph above almost all the probability changes occur between the values -2 and 2). We also need to include `bernoulli(link = "probit")` to ensure that `brm()` knows we want to run a probit model.

```{r}
#| cache: true
#| message: false
#| warning: false
fit_probit <- brm(
  bf(vote ~ 0 + issue + ussr_vote),
  prior = prior(normal(0, 2), class = b),
  family = bernoulli(link = "probit"),
  data = un |> filter(vote != 2), # Getting rid of abstentions
  cores = 4,
  chains = 4,
  backend = "cmdstanr",
  silent = 2,
  refresh = 0
)
```

Let's run a quick posterior-predictive check to make sure there are no glaring issues with our model. The $y$ line in the graph below shows the actual distribution of 1's and 0's in our data, and the $y_{rep}$ lines show different draws from the [posterior predictive distribution](https://mc-stan.org/bayesplot/reference/PPC-overview.html).

```{r}
pp_check(fit_probit, ndraws = 100)
```

Nice---the model is able to generally predict the 1's and 0's it was trained on.


```{r}
#| warning: false
#| include: false
fit_probit |> 
  gather_draws(`b_.*`, regex = TRUE) |> 
  # mutate(.value = pnorm(.value)) |>
  ggplot(aes(x = .value, y = .variable)) +
  stat_slabinterval(fill = "steelblue")
```

## Two Probit to Quit

Now that we're probit model experts, let's try something a little more interesting. Remember those abstentions we left out of the previous model? It's time to add them back in. 

There are multiple ways to construct a model with three possible outcomes (yes, no, abstain), but one of the most intuitive is to add a hurdle process to our original probit model. A hurdle model has two parts: first, we model whether some event will take place or not, then, conditional on the event taking place (i.e. overcoming the hurdle), we model the outcome of the event. So in our United Nations resolution example, the hurdle in question is whether a country takes any vote or abstains from voting. Then, if we predict that the country votes at all, we subsequently predict whether it will vote "yes" or "no" on the resolution.

We've covered how to model whether a country will vote "yes" or "no", but how do we model whether a country votes or abstains? The decision whether to vote or not to vote is itself a binary outcome---so we get to use the probit model once again! Two probits in one model! I'm still trying to come up with a catchy name for this type of hurdle probit model---ChatGPT was not much help:

> Q: What would be a catchy name for a hurdle probit model?

> A: How about "Trippy Hurdle Probit"? It combines the idea of a challenging hurdle with a humorous twist, suggesting that the model might encounter some unconventional obstacles along the way.

Unsurprisingly there is no "trippy hurdle probit" regression family in **brms**. Luckily **brms** allows you to create your own model families using the `custom_family()` function. The code for this section was adapted from these two great resources on custom families in **brms**: [Andrew Heiss's blog](https://www.andrewheiss.com/blog/2022/05/09/hurdle-lognormal-gaussian-brms/#hurdle-gaussian-model-with-a-custom-brms-family), and [brms package vignette](https://cran.r-project.org/web/packages/brms/vignettes/brms_customfamilies.html).

```{r}
hurdle_probit <- custom_family(
  "hurdle_probit",
  dpars = c("mu", "theta"),
  links = c("identity", "probit"),
  type = "int")
```

First we define a `custom_family()` with two distributional parameters, or `dpars`. The `mu` parameter corresponds to the yes/no part of the model and `theta` corresponds to the hurdle part. Unlike "mu" (which is required from **brms**), there is nothing special about choosing the label "theta" here. We're just following some [common notation](https://mc-stan.org/docs/2_20/stan-users-guide/zero-inflated-section.html).

```{r}
stan_funs <- "
  real hurdle_probit_lpmf(int y, real mu, real theta) {
    if (y == 2) {
      return bernoulli_lpmf(1 | theta);
    } else {
      return bernoulli_lpmf(0 | theta) +
             bernoulli_lpmf(y | Phi(mu));
    }
  }
"
stanvars <- stanvar(scode = stan_funs, block = "functions")
```

Next we need to write a new Stan function for **brms** to use. Note the conditional statement `y == 2` which corresponds to an abstention in our original data.

When building a new, and more complicated model like this it is a good idea to run a quick simulation to ensure things are working as expected. Our goal here is to determine whether the model can recover the same parameters we used to generate a synthetic data set. In the simulation code below, these parameters correspond to `x_coef` and `z_coef`.

```{r}
N <- 1000
x_coef <- -1
z_coef <- 2

hurdle_sim_data <- tibble(
  x = rnorm(N), # random variable affecting yes/no
  z = rnorm(N)  # random variable affecting hurdle
) |> 
  mutate(pr_abstain = pnorm(z * z_coef),              # probability of voting or abstaining
         abstain = rbinom(n(), 1, prob = pr_abstain), # binary abstain/vote
         pr_yes = pnorm(x * x_coef),                  # probability of yes or no
         yes = rbinom(n(), 1, prob = pr_yes),         # binary yes/no
         y = case_when(abstain == 1 ~ 2,              # final realized outcome
                       yes == 1 ~ 1,
                       yes == 0 ~ 0))
```

The synthetic data `hurdle_sim_data` can now be fed into a model using the custom family we created above.

```{r}
#| cache: true
fit_sim <- brm(
  bf(y ~ x,      # yes/no part
     theta ~ z), # hurdle part
  prior = prior(normal(0, 2), class = b, coef = x) +
          prior(normal(0, 2), class = b, coef = z, dpar = theta),
  data = hurdle_sim_data,
  family = hurdle_probit, # the custom_family we made
  stanvars = stanvars,    # the Stan function we made
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  silent = 2,
  refresh = 0
)
```

```{r}
fixef(fit_sim)
```

Taking a look at the coefficient estimates from our simulation model we see that we get roughly the same values as those that were used to generate the synthetic data! And just to further tie everything together, we can separate the two parts of the full hurdle model out and check the results. 

```{r}
# Using maximum likelihood here to save time
glm(abstain ~ z, 
    data = hurdle_sim_data, 
    family = binomial(link = "probit")) |> 
  broom::tidy()
```

Above we see the effect of `z` on whether an observation in our simulated data was marked "abstain" or not. Note that the coefficient estimate is the same as `theta_z` in the full hurdle model above.

```{r}
glm(yes ~ x, 
    data = hurdle_sim_data |> filter(abstain == 0), 
    family = binomial(link = "probit")) |> 
  broom::tidy()
```

And here we see the effect of `x` on the realized outcome `y` in the simulated data (after excluding the abstentions). Again, the coefficient estimate in this model nicely replicates the `x` estimate from the full hurdle model above.

Now that we're confident the hurdle probit model is working as intended, we can fit it to the real-world UN data we prepared earlier. We'll keep the part of the model predicting yes/no votes the same (i.e. using resolution issue and USSR vote variables as predictors). How should we predict whether the US votes or abstains from a particular resolution? Luckily for us, the **unvotes** data comes with a variable `importantvote`. Again, I'm no international relations expert but maybe if a resolution is deemed to be "important" the US will be more likely to weigh in---for or against.

```{r}
#| cache: true
#| message: false
#| warning: false
fit_hurdle <- brm(
  bf(vote ~ 0 + issue + ussr_vote, # yes/no model
     theta ~ importantvote),       # abstain/vote model
  family = hurdle_probit,
  stanvars = stanvars,
  data = un,
  cores = 4,
  chains = 4,
  backend = "cmdstanr",
  silent = 2,
  refresh = 0
)
```

Generating draws from the posterior predictive distribution from a custom family model in **brms** requires another user-defined function. [This vignette](https://cran.r-project.org/web/packages/brms/vignettes/brms_customfamilies.html) helps us out with this once again.

```{r}
posterior_predict_hurdle_probit <- function(i, prep, ...) {
  mu <- brms::get_dpar(prep, "mu", i = i)
  theta <- brms::get_dpar(prep, "theta", i = i)

  hu <- runif(prep$ndraws, 0, 1)
  ifelse(hu < theta, 2, rbinom(prep$ndraws, 1, pnorm(mu)))
}
```

Now we can check our results using `pp_check()`

```{r}
#| include: false
fit_hurdle |> gather_draws(`b_.*`, regex = TRUE) |> 
  # mutate(.value = pnorm(.value)) |> 
  ggplot(aes(x = .value, y = .variable)) +
  stat_slabinterval(fill = "steelblue")
```

```{r}
#| classes: preview-image
pp_check(fit_hurdle, ndraws = 100)
```

We see that the model does an okay job recovering the 0's (no), 1's (yes), and 2's (abstain) from the original data.

## Concluding Thoughts

If you're tackling a data problem involving binary outcomes and aiming to explicitly model abstentions, consider taking the trippy hurdle probit model for a spin. 

```{r}
sessionInfo()
```

